---
title: "EDA"

output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---


```{r defaults, echo = FALSE, message = FALSE, warning = FALSE}
# load necessary packages
library(tidyverse)
library(patchwork)
library(reshape2)
library(rstatix)
library(patchwork)
library(ggridges)
library(spatstat)
#library(ggpubr)
#library(GGally)
suppressMessages(library(dbplyr))
suppressMessages(library(energy))
suppressMessages(library(bookdown))
options(tibble.print_min = 5)
# set knitr defaults
knitr::opts_chunk$set(
    echo      = TRUE
  , message   = FALSE
  , fig.align = "center"
)
# set theme defaults
theme_set(
  theme_bw() +
  theme(
    legend.position = "bottom"
    , plot.title    = element_text(hjust = 0.5)
    , plot.subtitle = element_text(hjust = 0.5)    
    , plot.caption  = element_text(hjust = 0.0)
  )
)
# set color scale defaults
options(
    ggplot2.continuous.colour = "viridis"
  , ggplot2.continuous.fill   = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete   = scale_fill_viridis_d
```



```{r}
merged_data = 
  read_csv("data/Merged_expectation.csv", show_col_types = FALSE)

imputed_data = read_csv("data/Imputed_expectation.csv", show_col_types = FALSE)
```


# Time series plot
```{r}

merged_data %>% 
  group_by(year, `Developed / Developing Countries`) %>% 
  summarize(
    mean_val = mean(une_life)
  ) %>% 
  pivot_wider(
    values_from = mean_val,
    names_from = `Developed / Developing Countries`
  ) %>% 
  ggplot()+
  geom_line(aes(x=year, y=Developed, color='a'))+
  geom_line(aes(x=year, y=Developing, color='b'))+
  scale_color_manual(name = ' ', 
                     values =c("a"='black',"b"='red'), 
                     labels = c('Developed','Developing'))+
  geom_point(aes(x=year, y=Developed, color='a'),shape=19,size = 3)+
  geom_point(aes(x=year, y=Developing, color='b'),color = "red", shape=19,size = 3)+
  labs(title = sprintf("Time series plot for %s", "Life Expectancy") )+
  xlab("Year")+
  ylab(sprintf("%s","Life Expectancy"))
  
  
```






```{r}
Time_series_plot = function(variable,merged_data){
  merged_data %>% 
  group_by(year, `Developed / Developing Countries`) %>% 
  summarize_at(variable,mean,na.rm=TRUE) %>% 
  pivot_wider(
    values_from = variable,
    names_from = `Developed / Developing Countries`
  ) %>% 
  ggplot()+
  geom_line(aes(x=year, y=Developed, color='a'))+
  geom_line(aes(x=year, y=Developing, color='b'))+
  scale_color_manual(name = ' ', 
                     values =c("a"='black',"b"='red'), 
                     labels = c('Developed','Developing'))+
  geom_point(aes(x=year, y=Developed, color='a'),shape=19,size = 3)+
  geom_point(aes(x=year, y=Developing, color='b'),color = "red", shape=19,size = 3)+
  labs(title = sprintf("Time series plot for %s", variable) )+
  xlab("Year")+
  ylab(sprintf("%s",variable))
}


Time_series_plot("une_life",merged_data)

Time_series_plot("bmi",merged_data)
#Time_series_plot("bmi",imputed_data)
Time_series_plot("gni_capita",merged_data)

```




# Correlation:

## Pearson Correlation

Over the time:
```{r}

Pearson_Correlation = function(varnum, merged_data){
  
  correlation_result = tibble("date"=2000:2016, "Pearson_cor"=0,"Pearson_pval"=0)
  
  for (t in 2000:2016){
    merged_data_temp = 
      merged_data %>% 
      filter(year==t)
    
    correlation_result$Pearson_cor[t-1999] = 
    cor(merged_data_temp[,varnum], merged_data_temp$une_life, method = "pearson",use = "complete.obs")[1]
    
    correlation_result$Pearson_pval[t-1999] = 
    cor.test(merged_data_temp[,varnum][[1]], merged_data_temp$une_life, method = "pearson",use = "complete.obs")$p.value
    
  }
  
  return(correlation_result)
  
}

#Pearson_Correlation(6,merged_data)

Pearson_Correlation_result = tibble("date"=2000:2016)

colname = colnames(merged_data)
Pearson_Correlation_result %>%
  mutate(une_exp = Pearson_Correlation(6,merged_data)$Pearson_cor,
         une_exp_pval = Pearson_Correlation(6,merged_data)$Pearson_pval,
         alcohol = Pearson_Correlation(11,merged_data)$Pearson_cor,
         alcohol_pval = Pearson_Correlation(11,merged_data)$Pearson_pval
         )




```

## Distance Correlation

Distance correlation is a novel developed technique to discover the joint dependence of random variables developed by Székely et al. Unlike Pearson correlation, such Pairwise distance 
correlation is a distance method and therefore it can take into consideration of the non-linear relationship between two random variables, namely $X\in \mathbb{R}^p$ and $Y\in \mathbb{R}^q$ that can have arbitrary dimensions (p and q can be non-equal) to some extend. The correlation between two variables is said to be independent of distance correlation $\mathcal{R}(X,Y)=0$. 

In my article, denote the scalar product of two vectors $a$ and $b$ as <$a,b$> and conjugate of complex function f as $f$. Also define norm $|x|_p$ as Euclidean norm for $x\in\mathbb{R}^p$

<br>
<br>

### Definition 1
In weighted $l_2$ space, the $||\cdot||_w$-norm for any complex function $\zeta \in \mathbb{R}^p\times\mathbb{R}^q$ is defined by:
\[
\begin{equation}
    ||\zeta( a, b )||^2
_w =\int_{\mathbb{R}^{p+q}}|\zeta( a, b)|^2 w( a, b)\,d a\, d b
\end{equation}
\]
where $w(a,b)$ can be any arbitrary positive weighting function as long as the integral is well-defined. 

<br>
<br>


### Definition 2
Define the measure of dependence as:
\[
\begin{equation}
V^2(X,Y;w) = ||f_{X,Y}(a,b)-f_X(a)f_Y(b)||^2_w= \int_{\mathbb{R}^{p+q}}|f_{X,Y}(a,b)-f_X(a)f_Y(b)|^2 w( a, b)\,d a\, d b
\end{equation}
\]

#### Proof
*The criterion of independent means that for vectors $a$, $b$, the probability density function $f_X(a)$, $f_Y(b)$ and their joint probability density function $f_{X,Y}(a,b)$ satisfies $f_{X,Y}(a,b)=f_X(a)f_Y(b)$. Therefore, I have $V^2(X,Y;w)=||0||^2_w=0$ if they are independent hence the equation satisfy the condition that $V^2=0$ only if variables are independent.*



<br>
<br>

The choice of weighting function must satisfy invariance under transformations of random variables up to multiplication of a positive constant $\epsilon$: $(X,Y)\rightarrow (\epsilon X, \epsilon Y)$ and positiveness for any dependent variables. By the work of Szekely et. al., only non-integrable functions satisfy the condition for weight function. In order to specify the formula, I need the following lemma

<br>
<br>


### Lemma 1
For all x in $\mathbb{R}^d$:
\[
\begin{equation}
\int_{\mathbb{R}^d}\frac{1-cos<y,x> }{|y|^{d+1}_{d}}dy=c_d|x|=\frac{\pi ^{(1+d/2)}}{2 \Gamma ((d+1)/2)}|x|
\end{equation}
\]
where $\Gamma(\cdot)$ is the gamma function.


<br>
<br>

Now I can choose weight function as following:
\[
\begin{equation}
w(a,b) = (c_p c_q |a|
^{1+p}_p |b|
^{1+q}_q )^{-1}
\end{equation}
\]

By using such weight function and lemma 3.1, the definition of distance covariance will be well-defined. Note by the weight, I have $dw = (c_p c_q |a|
^{1+p}_p |b|
^{1+q}_q )^{-1}dadb$. This is proven in the following definition. \\

<br>
<br>

### Deinition 4
Define the distance covariance (dCov) for random vectors $X$ and $Y$ with finite expectation as the following non-negative figure:
\[
\begin{equation}
    V^2(X,Y) = ||f_{X,Y} (a,b)- f_X(a)f_Y (b)||^2
= \frac{1}{c_pc_q}\int_{\mathbb{R}^{p+q}}\frac{|f_{X,Y} (a,b) - f_X(a)f_Y (b)|^2}{|a|^{1+p}_p|b|^{1+q}_q}da db
\end{equation}
\]
where $a\in \mathbb{R}^p$, $b\in \mathbb{R}^q$ are two vectors that $a\in X$ and $b\in Y$\cite{szekely2007measuring}

<br>
<br>



```{theorem}
Here is my theorem.
```



```{r}
Distance_Correlation = function(varnum, merged_data){
  
  correlation_result = tibble("date"=2000:2016, "Distance_cor"=0,"Distance_pval"=0)
  
  for (t in 2000:2016){
    merged_data_temp = 
      merged_data %>% 
      filter(year==t) %>% 
      select(27,varnum) %>% 
      na.omit()
    
    correlation_result$Distance_cor[t-1999] = 
    dcor(merged_data_temp[,2], merged_data_temp$une_life,)[1]
    
    correlation_result$Distance_pval[t-1999] = 
    dcor.test(merged_data_temp[,2][[1]], merged_data_temp$une_life,R=1000)$p.value
    
  }
  
  return(correlation_result)
  
}



Distance_Correlation_result = tibble("date"=2000:2016)

colname = colnames(merged_data)
Distance_Correlation_result %>%
  mutate(une_exp = Distance_Correlation(6,merged_data)$Distance_cor,
         une_exp_pval = Distance_Correlation(6,merged_data)$Distance_pval,
         alcohol = Distance_Correlation(11,merged_data)$Distance_cor,
         alcohol_pval = Distance_Correlation(11,merged_data)$Distance_pval
         )



```







