---
title: "Data"

output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background Information of the datasets

<br>
<br>


### Life Expectancy Dataset

We choose WHO Life Expectancy Dataset as our main dataset in our study. The code is read as following:

```{r,message=FALSE, warning=FALSE}
suppressMessages(library(tidyverse))
suppressMessages(library(dbplyr))
options(tibble.print_min = 5)

#Get life expectancy data
LifeExpectancy = read_csv("data/who_life_exp.csv",col_types = cols(.default = col_guess())) %>% 
  janitor::clean_names()


```
Life expectancy datasets contains a total of ```r length(unique(LifeExpectancy$country))``` countries with year ranging from ```r min(unique(LifeExpectancy$year))``` to ```r max(unique(LifeExpectancy$year))```. There are a total of ```r length(LifeExpectancy)``` variables. Out of which, we consider the main 

*```r colnames(LifeExpectancy)[1]```: country name

*```r colnames(LifeExpectancy)[2]```: Three letter of the country id

*```r colnames(LifeExpectancy)[3]```: The region of the country

*```r colnames(LifeExpectancy)[4]```: year of all values stored

*```r colnames(LifeExpectancy)[5]```: Life expectancy at birth measured in year. It is continuous variable.

*```r colnames(LifeExpectancy)[6]```: Life expectancy at age 60 measured in year. It is continuous variable.

*```r colnames(LifeExpectancy)[7]```: Adult Mortality Rates of both sexes (probability of dying between 15 and 60 years per 1000 population)

*```r colnames(LifeExpectancy)[8]```: Death rate up to age 1

*```r colnames(LifeExpectancy)[9]```: Death rate between ages 1 and 4

*```r colnames(LifeExpectancy)[10]```: Alcohol, recorded per capita (15+) consumption (in litres of pure alcohol)

*```r colnames(LifeExpectancy)[11]```: Mean BMI (kg/$m^2$) (18+) (age-standardized estimate)

*```r colnames(LifeExpectancy)[12]```: Prevalence of thinness among children and adolescents. This is measured in a crude estimate percentage for children with BMI < (median - 2 s.d.)

*```r colnames(LifeExpectancy)[13]```: Prevalence of obesity  among children and adolescents. This is measured in a crude estimate percentage for children with BMI < (median - 2 s.d.)

*```r colnames(LifeExpectancy)[14]```: Hepatitis B (HepB) immunization coverage among 1-year-olds (%)

*```r colnames(LifeExpectancy)[15]```: Measles-containing-vaccine first-dose (MCV1) immunization coverage among 1-year-olds (%)

*```r colnames(LifeExpectancy)[16]```: Polio (Pol3) immunization coverage among 1-year-olds (%)

*```r colnames(LifeExpectancy)[17]```: Diphtheria tetanus toxoid and pertussis (DTP3) immunization coverage among 1-year-olds (%)

*```r colnames(LifeExpectancy)[18]```: Population using at least basic drinking-water services

*```r colnames(LifeExpectancy)[19]```: Number of medical doctors (per 10,000)

*```r colnames(LifeExpectancy)[20]```: Total density of hospitals per 100 000 population

*```r colnames(LifeExpectancy)[21]```: Gross national income per capita in dollars. This is measured from GHO server

*```r colnames(LifeExpectancy)[22]```: Domestic general government health expenditure (GGHE-D) as percentage of gross domestic product (GDP). This is measured from GHO server

*```r colnames(LifeExpectancy)[23]```: Current health expenditure (CHE) as percentage of gross domestic product (GDP) (%) 

*```r colnames(LifeExpectancy)[24]```: Population (thousands) 

*```r colnames(LifeExpectancy)[25]```: Mortality rate, infant (per 1,000 live births). This is measured from GHO server

*```r colnames(LifeExpectancy)[26]```: (Our response variable) Life expectancy at birth, total (years). This is measured from GHO server. It is contains less missing value than life expectancy. 

*```r colnames(LifeExpectancy)[27]```: Prevalence of HIV, total (% of population ages 15-49). This is measured from GHO server

*```r colnames(LifeExpectancy)[28]```: GNI per capita, PPP (current international $). This is measured from UNESCO  server

*```r colnames(LifeExpectancy)[29]```: Government expenditure on education as a percentage of GDP (%). This is measured from UNESCO  server

*```r colnames(LifeExpectancy)[30]```: Adult literacy rate, population 15+ years, both sexes (%). This is measured from UNESCO  server

*```r colnames(LifeExpectancy)[31]```: Mean years of schooling (ISCED 1 or higher), population 25+ years, both sexes. This is measured from UNESCO  server




<br>
<br>

### Country Code Dataset


Country code dataset contains the status of countries, such as the information of the independence status and its capital city. We include the variable specifies whether the country is a developed or developing country. The data is read in the following code and is merged by the code.  


```{r,message=FALSE, warning=FALSE}
country_code = 
  read_csv("data/country-codes.csv", show_col_types = FALSE) %>% 
  select(`ISO3166-1-Alpha-3`, `Developed / Developing Countries`)



merged_data_PM = merge(LifeExpectancy, country_code, by.x = "country_code", by.y = "ISO3166-1-Alpha-3" )
```


<br>
<br>


### PM2.5 dataset
PM2.5 dataset is a dataset that specifies the percentage of total population exposed to levels exceeding WHO guideline value. Such value exists over year 1960 to 2021 but most of the value before 2010 and after 2017 are missing.The following code read and clean the dataset.

merge the data with dataset contain pm 2.5 information:
```{r,message=FALSE, warning=FALSE}

#read pm2.5 data
PM_dataset = 
  read_csv("data/pm2.5.csv", show_col_types = FALSE, skip = 4) 

PM_dataset_clean = 
  PM_dataset %>% 
  janitor::clean_names() %>% 
  select(-(x1960:x1999),-(x2017:x2021)) %>% 
  pivot_longer(
    x2000:x2016,
    names_to = "year",
    names_prefix = "x",
    values_to = "PM_value"
  ) %>% 
  select(country_code, year, PM_value) %>% 
  mutate(year = as.numeric(year))
```

The following code is used to merge with all dataset.

```{r,message=FALSE, warning=FALSE}
#merged data with developed/ developing countries
merged_data_d = 
  merged_data_PM %>% 
  left_join(PM_dataset_clean, by = c("country_code","year"))
```


<br>
<br>

### Income level data

Income level data is a dataset specifies the income status. It is a factor variable with four levels, namely high income, low income, lower middle income and upper middle income. The dataset is cleaned and merged in the following code. 

```{r,message=FALSE, warning=FALSE}
#read income level data
Income_dataset = read_csv("data/income_level.csv", show_col_types = FALSE) %>% 
  janitor::clean_names() %>% 
  select(country_code,income_group)

#merged data with income level
merged_data_i = merged_data_d %>% 
  left_join( Income_dataset, by = c("country_code")) %>% 
  select(country_code:une_school,PM_value,income_group,`Developed / Developing Countries`)
```

<br>
<br>


### Health expenditure

Health expenditure dataset measures the health expenditure from government with respect to different countries. The dataset is measured in percentage of GDP or dollar value per capita. We choose percentage of GDP as our unit because the dataset measured in percentage of GDP has more complete values. Here is the coding for read, clean and merge the Health expenditure dataset to original dataset. 

```{r,message=FALSE, warning=FALSE}
health_exp = read_csv("data/government_compulsory_health expenditure_from_1970_2020.csv") %>% 
  janitor::clean_names() %>% 
  filter(measure=="PC_GDP") %>%
  filter(subject=="TOT") %>% 
  mutate("country_code"=location,
         "year" = time) %>% 
  select(country_code,year, value)

merged_data_h = 
  merged_data_i %>% 
  left_join(health_exp, by = c("country_code","year")) %>% 
  mutate("health_exp" = value) %>% 
  select(country_code:une_school,PM_value,income_group,health_exp,`Developed / Developing Countries`)
```

<br>
<br>


### Parliament Information

Parliament information data is a dataset that measured the percentage of women in national parliament around 2015~2018 for different countries. Since different countries have different election dates, and there is rare data before 2015, we choose the dataset that is around 2015~2018. Then, we adjust the date to year 2015, and add the percentage of women in lower house of national parliament to the last column of the original dataset. The following code cleans and merge the data:

```{r,message=FALSE, warning=FALSE}
women_in_parliament <-
  read_csv("data/around_2019_women_percent_in_national_parliaments.csv", skip = 5) %>% 
  janitor::clean_names() %>% 
  select(x2, elections_3, percent_w_6) %>% 
  mutate("country" = x2,
         "percent_w_in_lower_house" = percent_w_6,
         "year" = as.double(2015)) %>% 
  select(country, year, percent_w_in_lower_house)

merged_data = 
  merged_data_h %>% 
  left_join(women_in_parliament, by = c("country","year")) %>% 
  select(country_code:une_school,PM_value,health_exp,income_group,`Developed / Developing Countries`, percent_w_in_lower_house)

```


Store the data
```{r}
write.csv(merged_data , file = "data/Merged_expectation.csv")
```


# Impute

```{r,message=FALSE, warning=FALSE}
raw_mat = 
  merged_data %>% 
  select(-country, -country_code, -region,-income_group, -`Developed / Developing Countries`,percent_w_in_lower_house) %>% 
  as.matrix()

imputed_mat = filling::fill.KNNimpute(raw_mat, k = 5)

```

```{r,essage=FALSE, warning=FALSE}
#colnames_exp = colnames(LifeExpectancy)
Imputed_expectancy = merged_data
for (i in 1:(length(merged_data)-6)) {
  Imputed_expectancy[i+3][[1]] = imputed_mat[[1]][,i]
}
head(Imputed_expectancy)

write.csv(Imputed_expectancy , file = "data/Imputed_expectation.csv")
```







