---
title: "Linear Regression Model"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
```

After our EDA and statistical analysis, we consider to generate a linear regression to better express the relationship between Life Expectancy and other variables. In particular, we set une_life as our response variable and we only look at the year of 2015. 

First, we look at how many missing data we have on each column, at year 2015.

```{r, message=FALSE, warning=FALSE}
data <-
  read_csv("data/Merged_expectation.csv")
data %>% 
  filter(year == 2015) %>% 
  select(alcohol:percent_w_in_lower_house) %>% 
  is.na() %>% 
  colSums() %>% 
  knitr::kable(col.names = c("Counts of NA"))
```

Limited to the variable that we have less than 5% NA's (which is less than 10 NAs), also exclude variables that have high correlation value, we could generate our first basic linear model, with selected variables of alcohol, bmi, measles, basic_water, che_gdp (health expenditure per gdp), une_gni, PM_value, income_group, percent_w_in_lower_house, and our response variable of une_life.

* Model 1:
une_life ~ alcohol + bmi + measles + basic_water + che_gdp + une_gni + PM_value + income_group * percent_w_in_lower_house

Delete all the NA rows, consider intersection here, using step function with backwards direction. Then we generate our first model.


```{r,message=FALSE, warning=FALSE}
# delete all the Na columns
data_1 <- data %>% 
  filter(year == 2015) %>% 
  select(une_life, alcohol, bmi, measles, basic_water, che_gdp, une_gni, PM_value, income_group, percent_w_in_lower_house) %>% 
  drop_na()

m <- lm(une_life ~ alcohol + bmi + measles + basic_water + che_gdp + une_gni + PM_value + income_group * percent_w_in_lower_house, data = data_1)

m1 <- step(m, direction = "backward", trace = FALSE)
summary(m1)
```

From the above, variables like bmi, che_gdp, une_gni, PM_value are deleted according to backwards step function. One way to explain this is that these variables have high correlation with income_group--in higher income countries(une_gni), people have more access to healthy foods and healthy life styles (bmi), governments also spend more money on public health(che_gdp), as well as on building a clean environment(PM_value).

Another thing that needs to be aware of is that the intersection of income group and women in parliament (percent_w_in_lower_house). The variable percent_w_in_lower_house is significant itself, but it is more significant in middle income and upper income groups. In other words, after a country reaches a certain level of income, increasing women's voice in the country actually helps people live longer, which follows the intention that we chose to include this variable in the first place.

The model's R-adjusted score reaches 0.78.

Below are the analysis plots of the model. These plot shows that the residuals are normally distributed, independent, but may not have a constant variance--it has a linear trend in the end according to the third plot. In second plot, the points follows the line, so the regression function is linear. Therefore, maybe we don't need to add transformation to our response variable.

```{r,message=FALSE, warning=FALSE}
anova(m1)

par(mfrow=c(2,2))
plot(m1)
```

Second, we only look at the countries with existing une_literacy data and found out NA in other variables of these countries, at year 2015.

```{r,message=FALSE, warning=FALSE}
data %>% 
  filter(year == 2015) %>% 
  filter(is.na(une_literacy) == FALSE) %>% 
  select(alcohol:percent_w_in_lower_house) %>% 
  is.na() %>% 
  colSums() %>% 
  knitr::kable(col.names = c("Counts of NA"))
```

Here, limited to the variable that we have less than 5% NA's (which is less than 3 NAs), also exclude variables that have high correlation value, we could generate our second basic linear model, with selected variables of alcohol, bmi, measles, basic_water, che_gdp, une_gni, une_literacy, PM_value, income_group, percent_w_in_lower_house, and our response variable of une_life.

* Model 2:
une_life ~ alcohol + bmi + measles + basic_water + che_gdp + une_gni + une_literacy + PM_value, income_group + percent_w_in_lower_house + income_group * percent_w_in_lower_house

(Also consider intersection here)

```{r,message=FALSE, warning=FALSE}

```


* Model 3:
We could try all the variables, with imputed data. use step method to find the best method.
