---
title: "Linear Regression Model"
author: "Youlan Shen"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
```

First, we look at how many missing data we have on each column, at year 2015.

```{r, message=FALSE, warning=FALSE}
data <-
  read_csv("data/Merged_expectation.csv")
data %>% 
  filter(year == 2015) %>% 
  select(alcohol:percent_w_in_lower_house) %>% 
  is.na() %>% 
  colSums() %>% 
  knitr::kable(col.names = c("Counts of NA"))
```

Limited to the variable that we have less than 5% NA's (which is less than 10 NAs), also exclude variables that have high correlation value, we could generate our first basic linear model, with selected variables of alcohol, bmi, measles, basic_water, che_gdp (health expenditure per gdp), une_gni, PM_value, income_group, percent_w_in_lower_house, and our response variable of une_life.

* Model 1:
une_life ~ alcohol + bmi + measles + basic_water + che_gdp + une_gni + PM_value + income_group * percent_w_in_lower_house

Delete all the NA rows, consider intersection here, using step function with backwards direction. Then we generate our first model.


```{r}

```


Second, we only look at the countries with existing une_literacy data and found out NA in other variables of these countries, at year 2015.

```{r}
data %>% 
  filter(year == 2015) %>% 
  filter(is.na(une_literacy) == FALSE) %>% 
  select(alcohol:percent_w_in_lower_house) %>% 
  is.na() %>% 
  colSums() %>% 
  knitr::kable(col.names = c("Counts of NA"))
```

Here, limited to the variable that we have less than 5% NA's (which is less than 3 NAs), also exclude variables that have high correlation value, we could generate our second basic linear model, with selected variables of alcohol, bmi, measles, basic_water, che_gdp, une_gni, une_literacy, PM_value, income_group, percent_w_in_lower_house, and our response variable of une_life.

* Model 2:
une_life ~ alcohol + bmi + measles + basic_water + che_gdp + une_gni + une_literacy + PM_value, income_group + percent_w_in_lower_house + income_group * percent_w_in_lower_house

(Also consider intersection here)

```{r}

```


* Model 3:
We could try all the variables, with imputed data. use step method to find the best method.
