---
title: "World Map Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    source_code: embed
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)

library(plotly)
library(shiny)
```


```{r, include=FALSE}

df <- read_csv("./data/Merged_expectation.csv")
df <- df %>%
  select(country_code,country,une_life,year,region) 
  
```

```{r, include=FALSE}

mortality <- read_csv("./data/Merged_expectation.csv")
mortality <- mortality %>%
  select(country_code,country,adult_mortality,year) 
  
```

```{r, include=FALSE}

infant <- read_csv("./data/Merged_expectation.csv")
infant <- infant %>%
  select(country_code,country,infant_mort,year) 
  
```

```{r, include=FALSE}

alcohol <- read_csv("./data/Merged_expectation.csv")
alcohol <- alcohol %>%
  select(country_code,country,alcohol,year) 
  
```


```{r, include=FALSE}
bmi <- read_csv("./data/Merged_expectation.csv")
bmi <- bmi %>%
  select(country_code,country,bmi,year) 
  
```

```{r, include=FALSE}
che_gdp <- read_csv("./data/Merged_expectation.csv")
che_gdp <- che_gdp %>%
  select(country_code,country,che_gdp,year) 
  
```

Life Expectancy
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------


```{r}
year = df %>% distinct(year) %>% pull()

# selectInput widget
selectInput(
  inputId = "year", 
  label = h3("Select Year"),
  choices = year,
  selected = "2015")


```




Column {.tabset}
-----------------------------------------------------------------------

### Life Expectancy

```{r}
renderPlotly({
  # light grey boundaries
l <- list(color = toRGB("grey"), width = 0.5)

# specify map projection/options
g <- list(
  showframe = FALSE,
  showcoastlines = FALSE,
  projection = list(type = 'Mercator')
)

df %>% filter(year == input$year) %>% 
  plot_geo() %>% 
  add_trace(
  z = ~une_life, color = ~une_life, colors = 'Blues',
  text = ~country, locations = ~country_code, marker = list(line = l)
  ) %>% 
  colorbar(title = 'life expectancy') %>% 
  layout(
    title = str_c("life expectancy in ",input$year),
    geo = g
  )
})


```



### Description


This is the world map showing the life expectancy (years) of each country from 2000 to 2016. 

\ \par
Take the year of 2015 as an example,

\ \par
  * North America, Western and Northern Europe, Japan, New Zealand and Australia show longer life expectancy compared with other regions.
  
\ \par
  * Central and Southern Africa have relatively shorter life expectancy because the color is lighter,
  
  







Adult Mortality
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------


```{r}
year2 = df %>% distinct(year) %>% pull()

# selectInput widget
selectInput(
  inputId = "year2", 
  label = h3("Select Year"),
  choices = year2,
  selected = "2015")


```

Column {.tabset}
-----------------------------------------------------------------------

### Adult Mortality 

```{r}
renderPlotly({
  # light grey boundaries
l <- list(color = toRGB("grey"), width = 0.5)

# specify map projection/options
g <- list(
  showframe = FALSE,
  showcoastlines = FALSE,
  projection = list(type = 'Mercator')
)

mortality %>% filter(year == input$year2) %>% 
  plot_geo() %>% 
  add_trace(
  z = ~adult_mortality, color = ~adult_mortality, colors = 'Blues',
  text = ~country, locations = ~country_code, marker = list(line = l)
  ) %>% 
  colorbar(title = 'Adult Mortality ') %>% 
  layout(
    title = str_c("Adult Mortality in ",input$year2),
    geo = g
  )
})


```


### Description


This is the world map showing the adult mortality rate (probability of dying between 15 and 60 years per 1000 population) of each country from 2000 to 2016. 

\ \par
Take the year of 2015 as an example,

\ \par
  * Central and Southern Africa have relatively higher adult mortality rate.
  
\ \par
  * Lesotho, Central Africa Republic and Cote D'ivoire are the countries with the Top 3 highest adult mortality rate in the world and they are all in Africa.
  
\ \par
  * Syria has the highest adult mortality rate in Asia, which is probably because of the on-going war.










Infant Mortality
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------


```{r}
year3 = df %>% distinct(year) %>% pull()

# selectInput widget
selectInput(
  inputId = "year3", 
  label = h3("Select Year"),
  choices = year3,
  selected = "2015")


```

Column {.tabset}
-----------------------------------------------------------------------

### Infant Mortality 

```{r}
renderPlotly({
  # light grey boundaries
l <- list(color = toRGB("grey"), width = 0.5)

# specify map projection/options
g <- list(
  showframe = FALSE,
  showcoastlines = FALSE,
  projection = list(type = 'Mercator')
)

infant %>% filter(year == input$year3) %>% 
  plot_geo() %>% 
  add_trace(
  z = ~infant_mort, color = ~infant_mort, colors = 'Blues',
  text = ~country, locations = ~country_code, marker = list(line = l)
  ) %>% 
  colorbar(title = 'Infant Mortality ') %>% 
  layout(
    title = str_c("Infant Mortality in ",input$year3),
    geo = g
  )
})


```



### Description


This is the world map showing the infant mortality rate (death rate up to age 1) of each country from 2000 to 2016. 

\ \par
Take the year of 2015 as an example,

\ \par
  * Central and Southern Africa, Western and Southern Asia have relatively higher mortality rate, which is probably because of the underdeveloped medical level and poor hygiene.
  
\ \par
  * Sierra Leone, Central Africa Republic and Somalia are the countries with the Top 3 highest infant mortality rate in the world and they are all in Africa.
  
\ \par
  * Pakistan and Afganistan have the highest infant mortality rate in Asia.









Alcohol
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------

```{r}
year4 = df %>% distinct(year) %>% pull()

# selectInput widget
selectInput(
  inputId = "year4", 
  label = h3("Select Year"),
  choices = year4,
  selected = "2015")


```


Column {.tabset}
-----------------------------------------------------------------------

### Alcohol

```{r}
renderPlotly({
  # light grey boundaries
l <- list(color = toRGB("grey"), width = 0.5)

# specify map projection/options
g <- list(
  showframe = FALSE,
  showcoastlines = FALSE,
  projection = list(type = 'Mercator')
)

alcohol %>% filter(year == input$year4) %>% 
  plot_geo() %>% 
  add_trace(
  z = ~alcohol, color = ~alcohol, colors = 'Blues',
  text = ~country, locations = ~country_code, marker = list(line = l)
  ) %>% 
  colorbar(title = 'Alcohol') %>% 
  layout(
    title = str_c("Alcohol in ",input$year4),
    geo = g
  )
})


```


### Description


This is the world map showing the alcohol consumption level (liters of pure alcohol per capital of consumption) of each country from 2000 to 2016. 

\ \par
Take the year of 2015 as an example,

\ \par
  * Central and Northern Africa have the lowest alcohol consumption level in the world.
  
\ \par
  * Europe, North America, New Zealand and Australia have the highest alcohol consumption level in the world.
  
\ \par
  * Uganda has the highest alcohol consumption level in Africa.
  
  \ \par
  * Estonia, a country in the Northern Europe, has the highest alcohol consumption level in the world.












BMI
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------

```{r}
year5 = df %>% distinct(year) %>% pull()

# selectInput widget
selectInput(
  inputId = "year5", 
  label = h3("Select Year"),
  choices = year5,
  selected = "2015")


```


Column {.tabset}
-----------------------------------------------------------------------

### BMI

```{r}
renderPlotly({
  # light grey boundaries
l <- list(color = toRGB("grey"), width = 0.5)

# specify map projection/options
g <- list(
  showframe = FALSE,
  showcoastlines = FALSE,
  projection = list(type = 'Mercator')
)

bmi %>% filter(year == input$year5) %>% 
  plot_geo() %>% 
  add_trace(
  z = ~bmi, color = ~bmi, colors = 'Blues',
  text = ~country, locations = ~country_code, marker = list(line = l)
  ) %>% 
  colorbar(title = 'BMI') %>% 
  layout(
    title = str_c("BMI in ",input$year5),
    geo = g
  )
})


```



### Description


This is the world map showing the BMI level (age-standardized estimate) of each country from 2000 to 2016. 

\ \par
Take the year of 2015 as an example,

\ \par
  * America, Egypt, Saudi Arabia and Libya have the highest BMI level in the world.
  
\ \par
  * Central and Southern Africa, Eastern and Soutern Asia have a lower BMI level.
  
\ \par
  * Almost all European countries have a BMI level above 26.
  







Current Health Expenditure
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------

```{r}
year6 = df %>% distinct(year) %>% pull()

# selectInput widget
selectInput(
  inputId = "year6", 
  label = h3("Select Year"),
  choices = year6,
  selected = "2015")


```


Column {.tabset}
-----------------------------------------------------------------------

### Current Health Expenditure

```{r}
renderPlotly({
  # light grey boundaries
l <- list(color = toRGB("grey"), width = 0.5)

# specify map projection/options
g <- list(
  showframe = FALSE,
  showcoastlines = FALSE,
  projection = list(type = 'Mercator')
)

che_gdp %>% filter(year == input$year6) %>% 
  plot_geo() %>% 
  add_trace(
  z = ~che_gdp, color = ~che_gdp, colors = 'Blues',
  text = ~country, locations = ~country_code, marker = list(line = l)
  ) %>% 
  colorbar(title = 'Current Health Expenditure %') %>% 
  layout(
    title = str_c("Current Health Expenditure in ",input$year6),
    geo = g
  )
})


```


### Description


This is the world map showing the Current health expenditure (CHE) as percentage of gross domestic product (GDP) (%) of each country from 2000 to 2016. 

\ \par
Take the year of 2015 as an example,

\ \par
  * America and Sierra Leone have the highest current health expenditure in the world.
  
  \ \par
  * Generally speaking, Americas, Western and Northern Europe , Australia, New Zealand and Janpan have higher current health expenditure level.
  
  





